---
title: "Mini Project 1"
author: "Eden Ravecca"
date: "12/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Mini Project 1 using GBIF Prairie Falcon Data, NLCD Land Cover Data, USFS National Forest Climate Change Data, Tidycensus Cenus Data (by county), and TIGER/Line State shapefiles.  

### <span style="color: navy;"> For the mini project I am interested in exploring whether land cover, historical precipitation averages, and human population influence the liklihood of Prairie Falcon occurrence in Western US States. I downloaded Prairie Falcon data from GBIF, Land Cover data from NLCD, precipitation data from USFS National Forest Climate Change Maps, western states boundary shapefiles from TIGER/Line, and census data from Tidycensus.</span>

#### <span style="color: navy;"> Prairie Falcons are endemic to western North America and are typically found in deserts ranging from Southern Canada to Northern Mexico. Prairie Falcons are rarely seen east of the Mississippi River. They are more commonly found in arid climates, landscapes dominated by shrubs and lacking trees, and are a rare sighting in densely populated areas (unlike other falcons such as Peregrines and Kestrels). I am interested in exploring if their occurrence in the west is driven by low annual precipitation (historical average), land cover types, areas with low human population, or some combination of those variables.</span>    

#### <span style="color: navy;">To begin this project, I downloaded the relevant data from free internet sources. Due to the computational demands of using large spatial and temporal scales, I limited the data to only western states (from the Rocky Mountains to the pacific coast) and focused on a single year for falcon occurrence and land cover- 2016. The dependent variable is Prairie Falcon occurrence observations in 2016. The tabular data I am using includes: states (western US), date (year 2016), counties population. The spatial data includes: land cover 2016, climate (historical precipitation average), state polygons, county polygons.</span>    

```{r, load packages, warning=FALSE}
library(tidyverse)
library(sf)
library(terra)
library(tidycensus)
```

```{r, land cover data}
nlcd_data_terr <- terra::rast("/opt/data/MP/Ravecca_Eden_MP/NLCD/nlcd_2016_land_cover_l48_20210604.img") 
nlcd_data_terr2 <- terra::rast("/opt/data/MP/Ravecca_Eden_MP/NLCD/nlcd_2016_land_cover_l48_20210604.img") 
```

```{r, Prairie Falcon data}
falcon_csv <- read.csv("/opt/data/MP/Ravecca_Eden_MP/FalconData/PRFA_Data.csv")
states <- c( "Idaho", "Montana", "Wyoming", "Utah", "Nevada", "Arizona", "New Mexico", 
             "Colorado", "California", "Oregon", "Washington" ) # to filter states of interest at once
west_prfa <- falcon_csv %>% 
  dplyr::filter(., stateProvince %in% states) %>% # keep western states
  dplyr::select(., c(occurrenceID, locality, stateProvince, decimalLatitude, decimalLongitude, eventDate,
                     day, month, year)) %>% # keep relevant columns
  dplyr::filter(., year == "2016") %>% # keep only 2016 observations
  dplyr::rename(., c(state = stateProvince, lat = decimalLatitude, lon = decimalLongitude, date = eventDate, ID
                     = occurrenceID)) # change names to make sense
unique(west_prfa$year) # check
prfa.sf <- st_as_sf(west_prfa, coords = c("lon", "lat"), crs = 4326) # convert prfa gps points to sf object
prfa <- st_transform(prfa.sf, crs= crs(nlcd_data_terr))
```



```{r, states data}
states_data <- st_read("/opt/data/MP/Ravecca_Eden_MP/StatesPoly/tl_2012_us_state/tl_2012_us_state.shp")
w_states_shp <- states_data %>%
  dplyr::select(., c(STUSPS, NAME, geometry)) %>% # keep relevant columns
  dplyr::filter(., NAME %in% states) %>% # keep western states of interest
  dplyr::rename(., c(States = NAME, State_Abbreviation = STUSPS))
plot(w_states_shp, main= "Western US States") # check
all(st_is_valid(w_states_shp)) # make sure geometries are valid; TRUE
w_states_reproj <-  st_transform(w_states_shp, crs = crs(nlcd_data_terr)) # reproject states polys to match NLCD data
plot(nlcd_data_terr, main= "Land Cover Classes in US")
```

```{r}
plot(nlcd_data_terr)
plot(w_states_reproj, add = TRUE)
```


```{r, census data}
census_api_key("ccd5de8eb03adae16092bc5e1c6c40cb104f1e97", install = TRUE, overwrite = TRUE)
#readRenviron("~/.Renviron")
options(tigris_use_cache = TRUE)
reg.pop <- tidycensus::get_estimates(geography = "county",
                                     product = "population",
                                     state = c( "ID", "MT", "WY", "UT", "NV", "AZ", "NM", "CO", "OR", "WA", "CA" ), 
                                     year = 2016,
                                     key = key,
                                     geometry = TRUE,
                                     time_series = T) %>%
                                     st_transform(., st_crs(w_states_reproj))

w.cntypop <- filter(reg.pop, variable == "POP") # keep population value only (remove density)
unique(w.cntypop$variable)
head(w.cntypop)
all(st_is_valid(w.cntypop))
st_crs(w.cntypop) == st_crs(w_states_reproj) # check CRS matches
```

```{r, population summary}
cp.summary <- w.cntypop %>% 
  group_by(NAME) %>% 
  summarize(., avg_pop = mean(value)) %>%
  separate(NAME,c("county","state"),sep=",")
```

```{r, climate data}
precip_hist_terr <- terra::rast("/opt/data/MP/Ravecca_Eden_MP/NFCC_data/Ppt_annual_historical/Ppt_annual_historical.tif") 
x <- rast(extent= ext(nlcd_data_terr), crs= crs(nlcd_data_terr), resolution= 771)
ppt_hist_reproj <- terra::project(precip_hist_terr, x) # project to NLCD crs
plot(ppt_hist_reproj, main= "Historical Annual Precipitation Average in US")
```

```{r, crop rasters}
w.states.vect <- as(w_states_reproj, "SpatVector") # turn states shapefiles into SpatVector for cropping (in Terra)
plot(ppt_hist_reproj)
plot(w.states.vect, add= TRUE)
nlcd.fullres.crop <- terra::crop(nlcd_data_terr2, w.states.vect) # crop NLCD full res SpatRaster to States poly
ppt.crop <- terra::crop(ppt_hist_reproj, w.states.vect) # crop Precip. SpatRaster to States poly
```

```{r, plot cropped rasters}
plot(ppt.crop, main= "Historical Precipitation Average in Western US States")
plot(w.states.vect, add= TRUE)
```

```{r, check plots}
plot(nlcd.fullres.crop, main= "Land Cover Classes in Western US")
plot(w.states.vect, add= TRUE)
```

```{r, collapse=TRUE}
res(nlcd.fullres.crop) == res(ppt.crop)
crs(nlcd.fullres.crop) == crs(ppt.crop)
```

```{r, aggregate NLCD raster, collapse = TRUE}
nlcd.cats <- terra::cats(nlcd_data_terr2, layer=1) # define nlcd categories
nlcd_modeval <- terra::aggregate(nlcd.fullres.crop, fact=26, fun="modal") # aggregate to 780 m resolution and assign modal value from NLCD classes
print(nlcd_modeval)
plot(nlcd_modeval, type="classes")
```

```{r, resample NLCD raster, collapse = TRUE}
m <- rast(extent= ext(w.states.vect), crs= crs(nlcd.fullres.crop), resolution= 771) # blank raster for template
res(m)
crs(nlcd.fullres.crop) == crs(m)
ext(ppt.crop) == ext(m)
nlcd_771_mode <- terra::resample(nlcd_modeval, m, method= "near") # resample to 771 m resolution and assign NLCD value based on "nearest neighbor"
levels(nlcd_771_mode) <- nlcd.cats # nlcd Land Cover classes as levels
is.factor(nlcd_771_mode)
plot(nlcd_771_mode, type="classes", main= "Modal Land Cover Types in Western US States 771x771 km")
```

```{r, spatial join}
st_crs(prfa) == st_crs(cp.summary)
cp_prfa <- st_join(prfa, cp.summary, by = "state", left=TRUE)
cp_prfa.vect <- as(cp_prfa, "SpatVector") # turn county population & prfa data into SpatVector for rasterize (in Terra)
head(cp_prfa.vect)
```

```{r, extract}
ppt_values <- terra::extract(ppt.crop, cp_prfa.vect, fun= mean, method= "bilinear", na.rm=T)
lc_values <- terra::extract(nlcd_771_mode, cp_prfa.vect, fun= max, method= "simple", na.rm=T)
ppt_lc_val <- left_join(ppt_values, lc_values, by= "ID")
head(ppt_lc_val)
```



__Obviously you are working hard to get these done, but you didn't include the necessary number of tabular datasets and its not really clear what the relevant datasets are or why__


















